---
title: "2021 Microbial Community Analysis Workshop - Part1"
author: "UC Davis Bioinformatics Core"
output:
    html_document:
      keep_md: TRUE
---

# Reading in the experiment data and exploring the Phyloseq object

The phyloseq package has become a convenient way a managing micobial community data, filtering and visualizing data, performing analysis such as ordination. Along with the standard R environment and packages vegan and vegetarian you can perform virtually any analysis. Today we will

1. Load phyloseq object
2. Perform some QA/QC
3. Filter Data
5. Graphical Summaries
6. Ordination
7. Differential Abundances

## Load our libraries

```{r global_options}
# Set up global options for nice reports and keeping figures:
knitr::opts_chunk$set(fig.width=14, fig.height=8, fig.align="center",
                      warning=FALSE, message=FALSE)
```

Lets start by loading libraries

```{r libraries}
library(phyloseq)
library(phangorn)
library(ggplot2)

nice_colors = c("#999999", "#E69F00", "#56B4E9","#e98756","#c08160","#5800e6", "#CDDC49", "#C475D3", 
                "#E94B30", "#233F57", "#FEE659", "#A1CFDD", "#F4755E", "#D6F6F7","#EB6D58", "#6898BF")
```

## Read in the dataset, 

Lets first read in the Phyloseq object, this is the full dataset 197 generated from the same pipeline used in the data reduction section.

```{r readphyloseq}
load(file="phyloseq_nochim_silva.RData")
ls()

# set the data set used for analysis
ps = ps.silva.nochim

head(otu_table(ps))[,1:10]
head(sample_data(ps))
head(tax_table(ps))
head(refseq(ps))
head(rank_names(ps))
sample_variables(ps)
```

### Root the phylogenetic tree

Some analysis require the phylogenetic tree to be rooted. We use phanghorn root command to set our root.

```{r roottree}
set.seed(1)
is.rooted(phy_tree(ps))
phy_tree(ps) <- root(phy_tree(ps), sample(taxa_names(ps), 1), resolve.root = TRUE)
is.rooted(phy_tree(ps))
```

### The Phyloseq Object

A lot of information is in this object, spend somem time to get to know it.

* *Explore the sample data, how many Groups do we have? Temps? patterns of replicate?*
* *Using the `@` operator to access class objects are there any other info in the ps object not already looked at/*
* * Use `?"phyloseq-class"` to get information on the class and object.

```{r object}
ps
```

## Drawing our first plot
```{r firstplot, fig.height=7, fig.width=16, fig.align="center"}
ps
plot_bar(ps, fill = "Phylum") + theme(legend.position="bottom" ) +  scale_fill_manual(values = rainbow(length(unique(tax_table(ps)[,"Phylum"]))))
```

```{r get_next_rmd, eval=FALSE}
download.file("https://raw.githubusercontent.com/ucdavis-bioinformatics-training/2021-May-Microbial-Community-Analysis/master/data_analysis/mca-part2.Rmd", "mca-part2.Rmd")
```

```{r session_info}
sessionInfo()
```
